---
title: "R Notebook"
output: html_notebook
---

# Getting medical histories for participants

This notebook will walk you through getting medical histories for participants using the LabKey API in R. You are welcome to copy/paste any code from this notebook for your own scripts/notebooks.

## Contents:

* Import R libraries you need
* Helper function to access the LabKey API with R
* Getting medical history
  * Get details from a single `hes_*` table for a participant
  * Match accident and emergency episodes to overnight admissions
  * Find critical care periods within an overnight stay
  * Find hospital day appointments before or after another type of admission
  * Find details of diagnostic imaging within a time period
  * ECDS
* Mental health data
* Optional exercise

## Import R libraries you need

```{r}
library(tidyverse)
library(Rlabkey)
```

## Helper function to access the LabKey API with R

The following R function accesses the LabKey API, executes an SQL query to the specified database and returns it as a dataframe. We recommend putting this at the beginning of your scripts/notebooks and invoking it every time you want to access the database.

```{r}
labkey_to_df <- function(sql_query, database, maxrows){
    # Retrieve LabKey table based on a user SQL query.
    # 
    # Arguments:
    #       sql_query =   SQL query to ask the data base (character vector
    #                     of size 1).
    #       database  =   path containing the version of the programme
    #                     release to investigate (e.g.: "/main-programme/
    #                     main-programme_v14_2022-01-27")
    #       maxrows   =   number of rows to ask from the database. This 
    #                     parameter defaults to 100000000, but if the
    #                     output contains exactly the default value, then
    #                     we suggest increasing this value.
    
    labkey.setDefaults(baseUrl = "https://labkey-embassy.gel.zone/labkey/")
    
    labkey.executeSql(folderPath = database,
                      schemaName = "lists",
                      colNameOpt = "rname",
                      sql = sql_query,
                      maxRows = maxrows) %>%
        mutate(across(everything(), as.character))
}
```

To run my queries, I'll need to set up my database version:

```{r}
version <- "main-programme/main-programme_v17_2023-03-30"
```

## Getting medical history

### Get details from a single `hes_*` table for a participant

We're going to find all the diagnoses for an individual from the `hes_ae` table and the dates of those diagnoses. The participant we're looking at is ########.

```{r}
part_id <- "##########"
hes_sql <- paste("SELECT participant_id, arrivaldate, diag_all  
    FROM hes_ae
    WHERE participant_id = ", part_id, sep="")

hes_query <- labkey_to_df(hes_sql, version, 1000)
hes_query
```
The following code tidies up those diagnoses so that they're separated by commas and any empty fields are deleted:

```{r}
tidyup <- function(df){
  df[] <- lapply(df, gsub, pattern='\\|{2,}', replacement='')
  df[] <- lapply(df, gsub, pattern='\\|', replacement=', ')
  data.frame(df)
}
hes_query <- tidyup(hes_query)
hes_query
```
You can repeat a similar query with the other `hes_*` tables, pulling out other data. Please note that the dates in all the `hes_*` tables are named differently.

### Match accident and emergency episodes to overnight admissions

We're now going to combine this with a query of the `hes_apc` table. The following query finds all episodes where the participants went to accident and emergency, then were admitted for an overnight stay. It does this by checking for the `epikey` in both the `hes_ae` and `hes_apc` tables. We will find all the diagnoses and operations the participant received in both A&E and after admission.

```{r}
ae_apc_sql <- paste("SELECT ae.participant_id, 
  ae.arrivaldate as ae_arrival, 
  ae.diag_all as ae_diag, 
  apc.admidate as apc_arrival, 
  apc.disdate as apc_discharge, 
  apc.opertn_all as apc_operation, 
  apc.diag_all as apc_diag
    FROM hes_ae as ae, hes_apc as apc
    WHERE ae.epikey = apc.epikey
    AND ae.participant_id = ", part_id, sep = "")

ae_apc_query <- labkey_to_df(ae_apc_sql, version, 1000)
ae_apc_query <- tidyup(ae_apc_query)
ae_apc_query
```


### Find critical care periods within an overnight stay

This next query extends this further by checking the `hes_cc` table for all time spent in critical care while in admitted patient care. We can link the `hes_cc` table episodes to the `hes_apc` episodes using the `susrecid` column.

```{r}
ae_apc_cc_sql <- paste("SELECT ae.participant_id, 
  ae.arrivaldate as ae_arrival, 
  ae.diag_all as ae_diag, 
  apc.admidate as apc_arrival, 
  apc.disdate as apc_discharge, 
  apc.opertn_all as apc_operation, 
  apc.diag_all as apc_diag,
  cc.ccstartdate as cc_start,
  cc.ccdisdate as cc_discharge
    FROM hes_ae as ae
    FULL OUTER JOIN hes_apc as apc ON ae.epikey = apc.epikey 
    FULL OUTER JOIN hes_cc as cc ON apc.susrecid = cc.susrecid
    WHERE ae.participant_id = ", part_id,
  "ORDER BY ae.arrivaldate, apc.admidate", sep = "")

ae_apc_cc_query <- labkey_to_df(ae_apc_cc_sql, version, 1000)
ae_apc_cc_query <- tidyup(ae_apc_cc_query)
ae_apc_cc_query
```

### Find hospital day appointments before or after another type of admission

Now we're selecting a single row of the above table (in this case an arbitrary selection, but you may wish to select a row based on a diagnosis) and finding the discharge date. We will then find all outpatients appointments (day appointments at the hospital) that occurred in the 90 days after discharge. You could also search for appointment before a particular admission or procedure, or between certain events.

```{r}
date <- tail(ae_apc_cc_query$apc_discharge, n=1)
date_plus <- as.Date(date) + 90

op_sql <- paste("SELECT participant_id, 
  apptdate, tretspef, mainspef, diag_all, opertn_all
  FROM hes_op
  WHERE participant_id = ", part_id,
  " AND apptdate > '", date,
  "' AND apptdate < '", date_plus,
  "' ORDER BY apptdate", sep = "")

op_query <- labkey_to_df(op_sql, version, 1000)
op_query <- tidyup(op_query)
op_query
```
You may also wish to use `rttperstart` and `rttperend` to find the start and end of a series of appointments, but be aware that these columns are often blank and you will miss appointments by this method.

### Find details of diagnostic imaging within a time period

To find details of any diagnostic imaging, we will use the `did` table. There is no key that links entries in the `did` table to any of the `hes_*` tables, so we will use dates to link these.

In this query, we will use our query from above to filter out all admissions with a particular diagnosis (in this case R06, abnormalities of breathing) and find all did rows where the imaging date `did_date3` falls within the hospital stay.

```{r}
diag <- 'R06'

episodes <- ae_apc_cc_query %>% filter(grepl(diag, apc_diag))
row_counter <- 1
did_combined <- data.frame()

for(i in 1:nrow(episodes)) {
  episodes$did_index[i] <- row_counter
  did_sql <- paste("SELECT participant_id, 
    did_date3, ic_region_desc, ic_snomedct_desc, ic_sub_sys_desc, ic_nicip_desc
    FROM did
    WHERE participant_id = ", part_id, 
    " AND did_date3 >= '", episodes[i, "ae_arrival"],
    "' AND did_date3 <= '", episodes[i, "apc_discharge"],
    "'", sep = "")
  
  did_query <- labkey_to_df(did_sql, version, 1000)
  did_query$did_index <- row_counter
  did_combined <- rbind(did_combined, did_query)
  row_counter <- row_counter + 1
}

episodes <- merge(episodes, did_combined, by = c("did_index", "participant_id"))
episodes
```

###  ECDS

The ECDS table has different column headings, so we have to alter our queries accordingly. The following query gets all the diagnoses, treatments and investigations for a participant, with the date of arrival.

```{r}
ecds_sql <- paste("SELECT participant_id, seen_date, diagnosis_code_all, investigation_code_all, treatment_code_all ", 
    "FROM ecds ",
    "WHERE participant_id = ", part_id, sep="")

ecds_query <- labkey_to_df(ecds_sql, version, 1000)

ecds_query <- tidyup(ecds_query)
ecds_query
```
## Mental health data

Now we're going to look at the mental health tables. There are three sets of tables that cover different time periods:

* **mhmd** (Mental Health Minimum Dataset) – 2011-2014
* **mhldds** (Mental Health Learning Disabilities Dataset) – 2014-2016
* **mhsds** (Mental Health Services Dataset) – 2016 onwards

We're going to query these tables for information about the participant ############. `mhmd` and `mhldds` both follow similar schema, with `record`, `event` and `episode` tables, where a `record` is a superset of `episodes` which is a superset of `events`, linked by a `spell_id`. We'll start by querying the `mhmd` tables to find all details for our participant of interest.

```{r}
mh_part <- ###########

mhmd_sql <- paste("SELECT r.participant_id, r.mhd_mhmds_spell_id,
    ep.mhd_epitype, ep.mhd_epistart_date, ep.mhd_epiend_date,
    ev.mhd_eventtype, ev.mhd_eventdate
    FROM mhmd_v4_record as r
    LEFT JOIN mhmd_v4_episode as ep
        ON r.mhd_mhmds_spell_id = ep.mhd_mhmds_spell_id
    LEFT JOIN mhmd_v4_event as ev
        ON r.mhd_mhmds_spell_id = ev.mhd_mhmds_spell_id
    WHERE r.participant_id = ", mh_part, 
    sep = "")

mhmd_query <- labkey_to_df(mhmd_sql, version, 300000)
mhmd_query
```

We can use a very similar query with `mhldds`, as most of the columns are the same. You may find you need to change the column selections for certain data types.

```{r}
mhldds_sql <- paste("SELECT r.participant_id, r.mhd_mhmds_spell_id,
    ep.mhd_epitype, ep.mhd_epistart_date, ep.mhd_epiend_date,
    ev.mhd_eventtype, ev.mhd_eventdate
    FROM mhldds_record as r
    LEFT JOIN mhldds_episode as ep
        ON r.mhd_mhmds_spell_id = ep.mhd_mhmds_spell_id
    LEFT JOIN mhldds_event as ev
        ON r.mhd_mhmds_spell_id = ev.mhd_mhmds_spell_id
    WHERE r.participant_id = ", mh_part, 
    sep = "")

mhldds_query <- labkey_to_df(mhldds_sql, version, 50000)
mhldds_query
```

To query the more recent `mhsds` records we have to use a completely different method. We will query the curated tables, which are tied together by the `recordnumber`.

```{r}
mhsds_sql <- paste("SELECT p.participant_id, p.recordnumber,
    p.startdategmpregistration, p.enddategmpregistration,
    i.startdatewardstay, i.enddatewardstay, i.treatfunccodemh,
    c.refertotreatperiodstartdate, c.refertotreatperiodenddate,
    a.startdatecareclust, a.enddatecareclust
    FROM mhsds_curated_participant as p
    FULL OUTER JOIN mhsds_curated_inpatient as i
        ON p.recordnumber = i.recordnumber
    FULL OUTER JOIN mhsds_curated_community as c
        ON p.recordnumber = c.recordnumber
    FULL OUTER JOIN  mhsds_curated_assessment_diagnoses_and_cluster as a
        ON p.recordnumber = a.recordnumber
    WHERE p.participant_id = ", mh_part,
    sep = "")

mhsds_query <- labkey_to_df(mhsds_sql, version, 100)
mhsds_query
```
